---
title: "'Multi-Model Classification for CVD Prediction"
author: "Eliud Mule"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set Working Directory
```{r}
#setwd("C:/Users/ADMIN/OneDrive/Desktop/Eliud/R")
```



## **Task 1 : Data Import and Exploration

### **Load required libraries**
```{r}
library(tidyverse)      # Data Manipulation
library(caret)          # Unified ML Interface
library(pROC)           # ROC Analysis
library(e1071)          # SVM
library(class)          # KNN
library(randomForest)    # Random Forest
library(psych)
```

### **Import the dataset**
```{r}
data <- read.csv("medical_data.csv")
head(data)
```


### **Explore the dataset
```{r}
str(data)
describe(data)
```


### **Check for missing values**
```{r}
# Count missing values per column
missing_count <- colSums(is.na(data))
missing_count

```
```{r}

# Percentage of missing values
missing_percent <- (missing_count / nrow(data)) * 100
missing_percent

```


### **Visualize the distribution of the target variable CVD**
```{r}
ggplot(data, aes(x = CVD,fill = CVD)) +
  geom_bar(color = "black") +
  labs(title = "Distribution of CVD", x = "CVD Status", y = "Count") +
  theme_minimal()
```


### **Class Imbalance Ratio**
```{r}
# Count each class
class_counts <- table(data$CVD)

# Display counts
class_counts

# Calculate ratio
imbalance_ratio <- max(class_counts) / min(class_counts)

imbalance_ratio
```

## **Task 2 : Data Preprocessing**
### **Convert categorical variables to factors**
```{r}
data$gender <- as.factor(data$gender)
data$smoking <- as.factor(data$smoking)
data$alcohol <- as.factor(data$alcohol)
data$urban_residence <- as.factor(data$urban_residence)
data$family_history <- as.factor(data$family_history)
data$medication_adherence <- as.factor(data$medication_adherence)
```





## **Task 3 : Data Partitioning**
### **Split the dataset into training and testing sets**
```{r}
set.seed(2026)
train_index <- createDataPartition(data$CVD, p = 0.7, list = FALSE)
train_data <- data[train_index, ]
test_data <- data[-train_index, ]
```


```{r}
# Verify class distribution
cat("Training Set:\n")
print(table(train_data$CVD))
prop.table(table(train_data$CVD))
cat("\nTesting Set:\n")
print(table(test_data$CVD)) 
prop.table(table(test_data$CVD))
```


```{r}
preproc <- preProcess(train_data[, sapply(train_data, is.numeric)],
                      method = c("center", "scale"))

train_scaled <- predict(preproc, train_data)
test_scaled  <- predict(preproc, test_data)
```


## **Task 4 : Train Control Scheme**
```{r}
ctrl <- trainControl(
  method = "cv",              # 10-fold cross-validation
  number = 10,               # number of folds
  savePredictions = "final", # save predictions
  classProbs = TRUE,         # enable probabilities
  summaryFunction = twoClassSummary, # ROC, Sensitivity, Specificity
  verboseIter = FALSE        # suppress output
)
```


> Report :Cross-validation is used to evaluate model performance in a more reliable and robust manner. Instead of relying on a single train-test split, the data is divided into multiple folds, and the model is trained and validated across different subsets.
This reduces the risk of overfitting and ensures that the model generalizes well to unseen data.
The use of 10-fold cross-validation provides a good balance between computational efficiency and reliable performance estimation.


## **Task 5 :Model Training _ Logistic Regression**
```{r}
# Train Logistic Regression
log_model <- train(
  CVD ~.,
  data = train_data,
  method = "glm",
  family = binomial,
  trControl = ctrl
)
```



```{r}
# view results usingtab_model
library(sjPlot)
tab_model(log_model$finalModel,
          show.intercept = TRUE,
          show.se = TRUE,
          show.stat = TRUE,
          show.r2 = TRUE,
          show.ci = FALSE)

```

```{r} 
# Extract Coefficients
summary(log_model$finalModel)$coefficients
```

```{r}
# Obtain Odds Ratios
coefficients <- coef(log_model$finalModel)
odds_ratios <- exp(coefficients)

round(odds_ratios, 3)
```

## **Task 6 : Model Training - KNN**
```{r}
# Train KNN using scaled data
knn_model <- train(
  CVD ~ .,
  data = train_scaled,
  method = "knn",
  trControl = ctrl,
  tuneGrid = expand.grid(k = seq(1, 21, by = 2))
)


``` 

### **View the model summary**
```{r}
knn_model$results
```



### **View the best tuned model**
```{r}
knn_model$bestTune
```


### **Plot Accuracy vs K**
```{r}
plot(knn_model, main = "KNN: Accuracy vs k")
```



## **Task 7 : Model Training -SVM**
```{r}
# Training SVM using scaled data
svm_model <- train(
  CVD ~.,
  data = train_scaled,
  method = "svmRadial",
  trControl = ctrl,
  tuneGrid = expand.grid(
    C = c(0.25, 0.5, 1, 2, 4),
    sigma = c(0.01, 0.05, 0.1, 0.5)
  )
)
``` 


## **View the Model Summary**
```{r}
svm_model$results
```




### **View Optimal Parameters**
```{r}
svm_model$bestTune
```

### **Plot the Tuning Model**
```{r}
plot(svm_model, main = "SVM: ROC vs Parameters")
```



## **Task 8 : Model Training - Random Forest**
```{r}
rf_model <- train(
  CVD ~ .,
  data = train_data,
  method = "rf",
  trControl = ctrl,
  tuneGrid = expand.grid(mtry = c(2, 4, 6, 8)),
  ntree = 500
)
```


### **View the Model Summary**
```{r}
rf_model$results
```

### **View Optimal mtry**
```{r}
rf_model$bestTune
```



## **Variable Importance**
```{r}
importance <- varImp(rf_model)
print(importance)

plot(importance, main = "Random Forest: Variable Importance")
```



## **Task 9 : Model Evaluation and Comparison**
### **Predictions for all models**
```{r}
# Logistic Regression
log_pred  <- predict(log_model, test_data)
log_probs <- predict(log_model, test_data, type = "prob")

# KNN (scaled)
knn_pred  <- predict(knn_model, test_scaled)
knn_probs <- predict(knn_model, test_scaled, type = "prob")

# SVM (scaled)
svm_pred  <- predict(svm_model, test_scaled)
svm_probs <- predict(svm_model, test_scaled, type = "prob")

# Random Forest
rf_pred  <- predict(rf_model, test_data)
rf_probs <- predict(rf_model, test_data, type = "prob")
```


### **Confusion Matrices**
```{r}
# Convert test labels to factor with correct levels
test_data$CVD <- factor(test_data$CVD, levels = c("No", "Yes"))
test_scaled$CVD <- factor(test_scaled$CVD, levels = c("No", "Yes"))
```

```{r}
# Logistic
cm_log <- confusionMatrix(log_pred, test_data$CVD, positive = "Yes")
cm_log
```

```{r}
# KNN
cm_knn <- confusionMatrix(knn_pred, test_scaled$CVD, positive = "Yes")
cm_knn
```

```{r}
# SVM
cm_svm <- confusionMatrix(svm_pred, test_scaled$CVD, positive = "Yes")
cm_svm
```


```{r}
# Random Forest
cm_rf <- confusionMatrix(rf_pred, test_data$CVD, positive = "Yes")
cm_rf
```


### **ROC Curvess**
```{r}
log_roc <- roc(test_data$CVD, log_probs$Yes)
knn_roc <- roc(test_scaled$CVD, knn_probs$Yes)
svm_roc <- roc(test_scaled$CVD, svm_probs$Yes)
rf_roc  <- roc(test_data$CVD, rf_probs$Yes)

# Plot all
plot(log_roc, col = "blue", lwd = 2, main = "ROC Curve Comparison")
lines(knn_roc, col = "green", lwd = 2)
lines(svm_roc, col = "red", lwd = 2)
lines(rf_roc, col = "purple", lwd = 2)

legend("bottomright",
       legend = c("Logistic", "KNN", "SVM", "Random Forest"),
       col = c("blue", "green", "red", "purple"),
       lwd = 2)
```


### **AUC Comparison**
```{r}
auc_values <- data.frame(
  Model = c("Logistic", "KNN", "SVM", "Random Forest"),
  AUC = c(
    as.numeric(auc(log_roc)),
    as.numeric(auc(knn_roc)),
    as.numeric(auc(svm_roc)),
    as.numeric(auc(rf_roc))
  )
)

print(auc_values)
```













