---
title: "Logistic Regression for CVD Prediction"
author: "ELiud Mule"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set Working Directory 
```{r}
 #setwd("C:/Users/user/Desktop/Professional DS Mule/R")
```

## ** Task 1: Data Import and Exploration (10 Marks)**
### **Load libraries** 
```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(caret)
library(psych)
library(caret)      #for machine learning models
library(psych)      #for description of data
library(ggplot2)    #for data visualization
library(caretEnsemble) # enables creation of ensemble models
library(tidyverse)    #for data manipulation 
library(mlbench)     #for benchmarking ML models
library(flextable)     #to create and style tables
library(mltools)                
library(tictoc)       #for timing code execution
library(ROSE)          #for random sampling
library(ROCR)     #for ROC curve
library(smotefamily)     # for oversampling

```


### **Import the dataset**
```{r}
data <- read.csv("medical_data.csv")
head(data)


```


### **Explore the dataset**
```{r}
str(data)
describe(data)
```
 

### **Check for missing values per variable**
```{r}
data.frame(sapply(data, function(x) sum(is.na(x))))
```

```{r}
library(sjmisc)
frq(data, CVD)
```



### **Visualize the distribution of the target variable CVD**
```{r}
ggplot(data, aes(x = CVD,fill = CVD)) +
  geom_bar(color = "black") +
  labs(title = "Distribution of CVD", x = "CVD Status", y = "Count") +
  theme_minimal()
```

##  **Task 2: Data Preprocessing (15 Marks)**

### **Convert categorical variables to factors**
```{r}
data$gender <- as.factor(data$gender)
data$smoking <- as.factor(data$smoking)
data$alcohol <- as.factor(data$alcohol)
data$urban_residence <- as.factor(data$urban_residence)
data$family_history <- as.factor(data$family_history)
data$medication_adherence <- as.factor(data$medication_adherence)
```



## ** Task 3: Data Partitioning (10 Marks)**
### **Split the dataset into training and testing sets**
```{r}
set.seed(2026)
train_index <- createDataPartition(data$CVD, p = 0.7, list = FALSE)
train_data <- data[train_index, ]
test_data <- data[-train_index, ]
```


###  **Class distribution in both training sets**
```{r}
dim(train_data)
dim(test_data)

```




## ** Task 4: Model Training with trainControl (20 Marks)**
### **Set up trainControl for 10-fold cross-validation**
```{r}
train_control <- trainControl(method = "cv", number = 10, classProbs = TRUE, summaryFunction = twoClassSummary)

```

### **Train the logistic regression model**
```{r}
library(tictoc)
tic()
set.seed(123) # Set seed for reproducibility
model <- train(CVD ~ ., data = train_data,
        method = "glm",
        preProcess = c("scale","center"),
        trControl = train_control)
toc()
```


### **View Model**
```{r}
summary(model)
```


### **View the results above**
```{r}
library(sjPlot)
tab_model(model$finalModel,
          show.intercept = TRUE,
         # show.ci = FALSE,
          show.se = TRUE,
          show.stat = TRUE,
          show.r2 = TRUE)
```



## ** Task 5: Model Evaluation (25 Marks)**
### **Make predictions on the test set**
```{r}
pred_class <- predict(model,test_data)
pred_prob <- predict(model, test_data, type = "prob")

```


```{r}
test_data$CVD <- factor(test_data$CVD,levels = c("No", "Yes"))
pred_class <- factor(pred_class, levels = c("No", "Yes"))

```

### **Confusion Matrix**
```{r}
conf_matrix <- confusionMatrix(pred_class, test_data$CVD,positive = "Yes")
print(conf_matrix)

```


### **plot ROC curve and find area under the curve**
```{r}
library(pROC)
roc_obj <- roc(test_data$CVD, pred_prob[, "Yes"])
plot(roc_obj, col = "blue", main = "ROC Curve for CVD Prediction
",lwd = 2)
grid()
auc_value <- auc(roc_obj)
print(paste("AUC:", auc_value))

```
























