---
title: "Multi-Model Classification for Customer Churn Prediction"
author: "Eliud Mule"
date: '`r Sys.Date()`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set working Directory
```{r}
# setwd("C:/Users/user/Desktop/Professional DS Mule/R")
```


## **Task 1 : Data Import and Exploration

### **Load required libraries**
```{r}
library(tidyverse)      # Data Manipulation
library(caret)          # Unified ML Interface
library(pROC)           # ROC Analysis
library(e1071)          # SVM
library(class)          # KNN
library(randomForest)    # Random Forest
library(psych)
```

### **Load data from RData file**
```{r}
load("business_churn_data.RData")
ls()  # List objects in the environment to confirm data is loaded


```
```{r}
# Assuming the loaded data is in a variable named 'data'
data <- get(ls()[5])  # Get the sixth object in the environment, which should be our dataset
head(data)
```

```{r}
view(data)
```


### **Explore the dataset
```{r}
str(data)
describe(data)
```


### **Check for missing values**
```{r}
# Count missing values per column
missing_count <- colSums(is.na(data))
missing_count

```


### **Visualize the distribution of the target variable Churn**
```{r}
ggplot(data, aes(x = factor(Churn))) +
  geom_bar(fill = "steelblue") +
  labs(title = "Distribution of Churn", x = "Churn", y = "Count") +
  theme_minimal()
  
```

### **Class Imbalance Ratio**
```{r}
# Frequency table for Churn variable
churn_table <- table(data$Churn)
# Display the frequency table
churn_table
# Calculate the imbalance ratio
imbalance_ratio <- churn_table[1] / churn_table[2]
imbalance_ratio


```


## **Task 2 : Data Preprocessing**
### **Convert target variable to factor**
```{r}
data$Churn <- factor(data$Churn, levels = c(0, 1), labels = c("No", "Yes"))

```

### **Convert Categorical variables to factors**
```{r}
categorical_vars <- c("Gender","ContractType","PaymentMethod","InternetService","TechSupport","OnlineSecurity","LoyaltyProgram","StreamingTV","StreamingMovies") 
data[categorical_vars] <- lapply(data[categorical_vars], as.factor)
```



## **Task 3 : Data Partitioning**
### **Split the dataset into training and testing sets**
```{r}
set.seed(2026)  # For reproducibility
#Stratified sampling to maintain class distribution
train_index <- createDataPartition(data$Churn, p = 0.7, list = FALSE)
train_data <- data[train_index, ]
test_data <- data[-train_index, ]

```


### **Check class distribution in training and testing sets**
```{r}
table(train_data$Churn)
table(test_data$Churn)

```

### **Apply scaling to numeric features**
```{r}
numeric_vars <- sapply(train_data, is.numeric)
prepocessing_model <- preProcess(train_data[, numeric_vars], method = c("center", "scale"))
train_data_scaled <- predict(prepocessing_model, train_data)
test_data_scaled <- predict(prepocessing_model, test_data)

```


## **Task 4 : Train Control Scheme**
### **Define train control for cross-validation**
```{r}
ctrl <- trainControl(
  method = "cv",          # Cross-validation
  number = 10,           # Number of folds
  classProbs = TRUE,     # Needed for AUC calculation
  summaryFunction = twoClassSummary,  #ROC,Sensitivity,Specificity
  savePredictions = "final" # Save predictions for the best model
)
```



### **Task 5 :Model Training - Logistic Regression**
```{r}
set.seed(2026)
logistic_model <- train(Churn ~ ., data = train_data, method = "glm", family = "binomial", trControl = ctrl, metric = "ROC")


```

```{r}
# view coeeficients of the logistic regression model
summary(logistic_model$finalModel)$coefficients
```


```{r}
# Calculate odds ratios using tab_model function from sjPlot package
library(sjPlot)
tab_model(logistic_model$finalModel, show.ci = FALSE, show.se = FALSE, show.stat = FALSE, show.p = FALSE, show.aic = FALSE)
```


## **Task 6 : Model Training -KNN**
```{r}
set.seed(2026)
knn_model <- train(Churn ~ ., data = train_data_scaled, method = "knn", trControl = ctrl, metric = "ROC",
                   tuneGrid = expand.grid(k = seq(1, 21, by = 2)))  
```


```{r}
# View the optimal k value for KNN
knn_model$bestTune
```


```{r}
# Plot KNN performance across different k values
plot(knn_model,main = "KNN : Accuracy vs K")
```

..
## **Task 7 : Model Training - SVM**
```{r}
set.seed(2026)
svm_model <- train(Churn ~ ., data = train_data_scaled, method = "svmRadial", trControl = ctrl, metric = "ROC",
                   tuneGrid = expand.grid(C = c(0.25, 0.5, 1,2,4), sigma = c(0.01,0.05, 0.1, 0.5)))

```

```{r}
# View the best parameters for SVM
svm_model$bestTune
```


```{r}
#Plot Tuning results for SVM
plot(svm_model, main = "SVM : ROC vs Parameters")
```


## **Task 8 : Model Training - Random Forest**
```{r}
rf_model <- train(Churn ~ ., data = train_data, method = "rf",  trControl = ctrl, metric = "ROC",
                   tuneGrid = expand.grid(mtry = c(2, 4, 6, 8, 10)))
```


```{r}
# View the best mtry value for Random Forest
rf_model$bestTune
```



```{r} 
#Plot variable importance for Random Forest
varImp(rf_model)
plot(varImp(rf_model), main = "Random Forest : Variable Importance")

```


## **Task 9 : Model Evaluation and Comparison**
### **Predictions for all models**
```{r}
logistic_preds <- predict(logistic_model, test_data, type = "prob")[, "Yes"]
knn_preds <- predict(knn_model, test_data_scaled, type = "prob")[, "Yes"]
svm_preds <- predict(svm_model, test_data_scaled, type = "prob")[, "Yes"]
rf_preds <- predict(rf_model, test_data, type = "prob")[, "Yes"]
```


### **Confusion Matrices for each Model**
```{r}
# Confusion Matrix for Logistic Regression
logistic_cm <- confusionMatrix(predict(logistic_model, test_data), test_data$Churn,positive = "Yes")
logistic_cm
```

```{r}
# Confusion Matrix for KNN
knn_cm <- confusionMatrix(predict(knn_model, test_data_scaled), test_data$Churn,positive = "Yes")
knn_cm
```
```{r}
# Confusion Matrix for SVM
svm_cm <- confusionMatrix(predict(svm_model, test_data_scaled), test_data$Churn,positive = "Yes")
svm_cm
```


```{r}
# Confusion Matrix for Random Forest
rf_cm <- confusionMatrix(predict(rf_model, test_data), test_data$Churn,positive = "Yes")
rf_cm
```

### **ROC Curves and AUC for each Model**
```{r}
# All ROC curves in one plot
roc_logistic <- roc(test_data$Churn, logistic_preds)
roc_knn <- roc(test_data$Churn, knn_preds)
roc_svm <- roc(test_data$Churn, svm_preds)
roc_rf <- roc(test_data$Churn, rf_preds)
plot(roc_logistic, col = "blue", main = "ROC Curves for All Models")
plot(roc_knn, col = "red", add = TRUE)
plot(roc_svm, col = "green", add = TRUE)
plot(roc_rf, col = "purple", add = TRUE)
legend("bottomright", legend = c("Logistic Regression", "KNN", "SVM", "Random Forest"), col = c("blue", "red", "green", "purple"), lwd = 2)

```



### **AUC Comparison**
```{r}
auc_values <- data.frame(
  Model = c("Logistic Regression", "KNN", "SVM", "Random Forest"),
  AUC = c(auc(roc_logistic), auc(roc_knn), auc(roc_svm), auc(roc_rf))
)
print(auc_values)
```
















































